---
title: "Final Report: Dietary Yogurt Influence on Human Fitness and Disease"
author: "Brian Rezende"
date: "December 5, 2018"
output:
github_document: default
bibliography: references.bib
csl: bioinformatics.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

One of the most essential microbiome communities in humans is the gut microbiome that is enclosed by the gastrointestinal tract (GI). Microbiomes are large operating communities that are composed of about 38 trillion microbiome cells and 30 trillion human cells, thus accounting for every surface of the human body [@Sender2016]. Within the GI, these numbers will fluctuate as a result of long-term or short-term resident bacteria that is acquired through diet [@deAlmada2015]. Rather than finding trillions of bacterial cells, the GI contains approximately 1011-1012 bacteria cells per gram of luminal content [@Rastall2004]. Though this is a drastic subset of the entire human microbiome community, this region still exhibits microbial density and diversity that increases distally all the way through and towards the colon. The human gut is an interesting region to study as it contains microbiota that carry out vital functions for humans and can dictate our health. Previous work has demonstrated that the human gut is primarily dominated by *Phylums Bacteroidetes* and *Firmicutes* that contain *Genus* such as *Bifidobacteria*, *Lactobacillus*, *Streptococcus*, and *Clostridia*, while other types of bacteria are seen yet found in lesser proportions [@Sekirov2010]. Though the exact composition of the mammalian GI is host specific, one way we may be able to reach equilibrium and quantify changes in the GI is through the use of probiotics.

Probiotics have gained increased popularity and attention through the use of advertisement and commercials, which is supported by its ability to prevent multiple gastrointestinal diseases, coupled with various host health benefits seen by modulating immunological responses. For instance, a study has shown that yogurt, milk, and cheese consumption is associated with decreased common carotid artery intima-media thickness **(CCA-IMT)** and is able to reduce cardiovascular disease risk factors [@Ivey2011]. Yogurt consumption in specific was shown to decrease CCA-IMT in the healthy subjects as opposed to the other individuals who had lower consumption of the dairy products. Other studies indicate that frequent consumption of yogurt intake also aids in lowering diabetes risk, enhancing development of host immunity, and lowering the risk for dysbiosis and chronic kidney disease. 
The most traditional commensal microbiota that are consumed for their potential health benefits include *Bifidobacterium* and *Lactobacillus*. *Lactobacilli*-containing yogurt in particular has been shown to alter the composition of the gut microbiota of healthy individuals and it reports changes for two groups of bacteria: *Bacteriodes* and *Prevotella*, along with the *C. coccoides-E. Rectalei* group [@UYENO2008]. Though probiotics have the ability to increase the gut content of certain strains, they do not significantly alter the microbial community structure, which is key for maintaining proper function of the human gut. 

Another impressive feat of probiotic yogurt consumption is seen in individuals infected by *Helicobacter pylori*. This infection results in chronic gut inflammation and causes an imbalance of *Bifidobacterium/E.coli* ratio. This imbalance was found to be rectified by yogurt consumption as it restores an optimal ratio coupled with altering the serum immune response [@Lisko2017]. Probiotics accomplish this by increasing the gut microbiota to be *Bifidobacterium* dominant, which kids that suffer from *H. pylori* are deficient in, by stimulating a subtle inflammation in the intestine that elevates secretion of mucosal immunoglobulin A (IgA) into the serum [@Yang2012]. These findings help to confirm that yogurt consumption can modulate both cellular and humoral immunity in children. It is thus no surprise as to why yogurt is supposed to be an integral part of our diet and is seen as a popular health-promotion food. 

An important relationship between host and gut, is the gut-brain axis. This axis allows for communication between the gut and host as it is a bidirectional neurohumoral communication system that is able to dictate behavior and health. Signals from the brain are able to modify motor, sensory, and secretory modalities of the GI, while signals from the gut affect emotional behavior, stress, and pain systems through neural, endocrine, and immune pathways [@KatoKataoka2016]. Probiotic consumption may play a role by modulating this communication system which affects emotion in health and disease and positively benefits the host. Probiotics accomplish this by maintaining the mucosal barrier function under stressful situations and by mitigating stress induced glucocorticoid and inflammatory cytokine responses [@KatoKataoka2016]. They also possess the ability to reduce the mRNA expression of GABA receptors in the brain by modulating the gut-brain axis via vagal pathways [@Bravo2011]. This amalgamation of these processes thus exhibit one of the direct pathways of probiotics on human fitness. 

A particular genus of interest within probiotics is *Lactobacillus* as they demonstrate a stronger presence over *Bifidobacteria* species within the host-gut system. Though this genus makes up only **1%** of the total microbatercial community, it is still comprised of many diverse organisms. Current work demonstrates that it includes over 25 species that are capable of inhibiting various habitats such as GI tracts to other fermented foods [@Heilig2002]. These species are what influence health, behavior, and overall structure of the gut microbiota, making it particularly interesting to explore and track as consuming yogurt containing these strains increase their levels found within individuals throughout aging. Identified species of *Lactobacillus* are vastly malleable to other pathways in the body and can possibly serve as therapeutics for an array of diseases and biological dysfunctions in the future. 

Further research must still be done in order to determine the direct mechanisms that probiotics operate and to reveal species that are still undiscovered. Questions remain on whether or not probiotics play a direct or indirect effect in the pathways that they are involved in. One of the pitfalls in being able to answer such questions is that studies require longer time periods in order to better determine the effects of consumption and then quantify the changes it brings about. 
Given the diverse relationships that probiotics may affect, especially in the GI tract, this analysis will aim to analyze the effects of prolonged probiotic intake in the modulation of the GI microbiome. Since bacteria has the ability to act in positive ways and influence health and behavior, I will focus on *Lactobacilli* and *Bifidobacteria* populations and how they change over time. Given their inherent change, I hypothesize that changes in bacterial composition within the GI tract are essential for increasing an individual's fitness, specifically in combating stress and cardiovascular diseases.

# Methods

## Study design

Lisko and his contributors utilized a multi-approach that combined various methods of sequencing in order to explicate bacterial communities and *Lactobacilli* and *Bifidobacteria* populations within healthy individuals. These individuals were utilized to investigate the effects of yogurt consumption on the GI microbiome bacteria community composition over the span of 42 days. Lisko et al required no dietary or exercise logs to be kept, coupled with allowing subjects to maintain their regular diet and continue with their activities throughout the study [@Lisko2017]. The yogurt of choice was determined through viability tests and serial dilutions that were used to enumerate colony forming units (CFUs), thus the yogurt with the highest CFUs per gram for both *Lactobacilli* and *Bifidobacteria* were selected for the study. Lisko et al., ultimately aimed to determine microbial community structure after a high dosage of yogurt was fed to 8 healthy human subjects. They performed their analysis in a noninvasive manner and were eventually able to utilize three different molecular methods. 

## Sample origin and sequencing

Daniel Lisko, Patricia Johnston, and Carol Johnston of Youngstown State University began the study by recruiting 8 healthy subject volunteers (2 females and 6 males)  between the ages of 18 and 54. Participants had no prior history of lactose intolerance and were barred from taking antibiotics six months prior to the study and throughout the study. In order to study the effects of dietary yogurt on the healthy human GI microbiome, participants were required to consume **250 g** of a selected yogurt per day for 42 consecutive days. Following this consumption, fecal samples were collected at 7 different time points and placed through various methods of sequencing. From the fecal matter of each subject, a single DNA extraction was performed and terminal restriction fragment length polymorphism (T-RFLP) analysis was used. Lisko et al. chose this fingerprinting technique in order to monitor changes in community structure. This type of technique is useful in being able to explore microbial diversity of predominant populations and provides higher throughput than other cultivation-dependent methods [Schtte2008]. This type of analysis then amplified the bacterial rDNA using selective primers, which were used for PCR reactions. Only 3 of the 8 subjects DNA extractions underwent cloning and Sanger sequencing in order to obtain 16s rDNA gene sequences for *Lactobacilli* and *Bifidiobacteria* phylogenetics using the Lac 1, lac 2, G-Bifid-F and G-Bifid-R primers. As a result, 520 clones (273 from Lactobacillus and 243 from Bifidobacteria) were Sanger sequenced [@Lisko2017]. Next, the genes found at the beginning and end of the study from each bacteria underwent qPCR reactions. This then allowed Lisko et al. to utilize NMDS to compare T-RFLP restriction fragments (T-RFs) across 5 subjects and the 1 control subject. Lastly, Next Generation Sequencing (NGS) was used to amplify the original 16s rDNA from the test subjects at different time points using total DNA. An Illumina MiSeq platform gave them the opportunity to use paired-end sequencing, which were then joined and removed depending on various quality checks. Lisko and his colleagues then clustered the quality filtered sequences into operational taxonomic units (OTU). The most abundant sequences from each taxonomic unit were used to represent the entire OTU, which were then assigned taxonomy. Lastly, alpha and beta diversity were established using rarefied sequences using Chao1 and Shannon Diversity and weighted UniFrac distances from unique branch lengths in pairs of communities for the phylogenetic tree. 

## Computational

For this analysis I downloaded the raw sequence data from the NCBI Sequence Read Archive under the study PRJNA351747 with accession numbers SAMN05990293-990309. A bash script was created to perform the download of the data, with a pipeline that placed the sequences into a metadata folder. I then utilized fastq dump to convert the files from their proprietary form to fastq files. Using the metadata produced from Lisko’s initial study, I ran the series of fastq files generated through a pipeline of quality checks followed by filtering, trimming, and cleaning completed using a DADA pipeline. DADA is a program that is used as a universal tool of being able to correct errors and collapse sequences that are similar onto one another. DADA parameters were set to locally align the number of bands and to ensure that biological variation is not mistaken as sequencing errors. Rather than producing operational taxonomic units (OTUs) this method will produce amplicon sequence variants (ASVs) that are trimmed and denoised using the filterandTrim function [@callahan2016]. We then cleaned up the outputted DATA material even further to remove chimeras and accurately assign taxonomic labels to each sequence variant. Once taxonomy was assigned to our sequence variants, a conversion from sequences to fasta files were necessary in order to plot out phylogeny. The sequence names remained as the labels for the sequence and a taxonomic table was generated. The phylogeny was read in from the fasta generated and a phyloseq object was constructed straight from the DADA outputs. We mutated the resulting metadata to separate a column that condensed 3 variables into one, thus creating a second metadata table that included time points, treatment, and the different subjects. In order to perform our desired analysis the phyloseq was converted into a datatable, then melted to be used for dpylr and ggplot2 graphing functions. A series of phyloseq graphing tools were implemented in order to delve deeper into the spread of taxa within samples and their respective abundance within phylums. Other phyloseq tools and graph manipulations utilized were subsets of data that were filtered and trimmed [@phyloseq2013]. Using the revised metadata and their fastq files we created the following figures to infer further curiosities about the bacterial communities found in yogurt probiotics,  specifically in the genus of both *Lactobacillus* and *Bifidobacteria*. 

# Results

```{r load-packages-and-data, echo=FALSE, message=FALSE, warning=FALSE}
# load packages as needed
library("dplyr")
library("tidyr")
library("citr")
library("knitr")
library("ggplot2")
library("phyloseq")

# load data produced from analysis scripts using
# something like load("output/processed_data.Rdata")
load("output/phyloseq_obj.R")
load("output/melted_obj.R")
```

```{r most-abundant-genus-day-0, echo=FALSE, message=FALSE, warning=FALSE}
melted_phyloseq %>%
  filter(timepoints == "0") %>%
  filter(!is.na(Genus)) %>%
  group_by("Genus") %>%
  ggplot(aes(x = Genus,
             y = Abundance,
             fill = Subject))+
  geom_bar(stat = "Identity")+
  xlab("Genus")+
  ylab("Abundance")+
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1))+
  ggtitle("Abundance of Bacteria on Day 0 per Subject")
  
```

**Figure 1:** Bacterial community composition in GI tract for each of the 8 healthy subjects on day 0. Figure shows genus species levels without any affect of yogurt consumption. *Bifidobacterium* demonstrates greater abundance of species over *Lactobacillus*. Subject G demonstrates no presence of *Bifidobacterium*. Though it is *very* small (pertaining to Subject C and D), all subjects show traces of the species *Lactobacillus*. Subject E is missing from from this data collection date. 

```{r most-abundant-genus-day-7, echo=FALSE, warning=FALSE}
melted_phyloseq %>%
  filter(timepoints == "7") %>%
  filter(!is.na(Genus)) %>%
  group_by("Genus") %>%
  ggplot(aes(x = Genus,
             y = Abundance,
             fill = Subject))+
  geom_bar(stat = "Identity")+
  xlab("Genus")+
  ylab("Abundance")+
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1))+
  ggtitle("Abundance of Bacteria on Day 7 per Subject")

```

**Figure 2:** Bacterial community composition in GI tract for each of the 8 healthy subjects on day 7. Figure shows an increase in some species such as *Shingella* and *Salmonella*, while presence of *Lactobacillus* is vastly decreased in subject B. On day 7, subject E appears and demonstrates that it contains *Bifidobacterium*. In contrast, subjects A, E, and G show an increase in both genus species. Subjects C and F appear to be missing from this data collection date. 

```{r most-abundant-genus-day-42, echo=FALSE, warning=FALSE}
melted_phyloseq %>%
  filter(timepoints == "42") %>%
  filter(!is.na(Genus)) %>%
  group_by("Genus") %>%
  ggplot(aes(x = Genus,
             y = Abundance,
             fill = Subject))+
  geom_bar(stat = "Identity")+
  xlab("Genus")+
  ylab("Abundance")+
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1))+
  ggtitle("Abundance of Bacteria on Day 42 per Subject")
```
 
**Figure 3:** Bacterial community composition in GI tract for each of the 8 healthy subjects on day 42. Figure shows increased abundance of the two popular species, *Bifidobacterium* and *Lactobacillus*, which indicates they're almost reaching equilbirium. On day 42 the presence of these genuses fluctuate further compared to previous time points and almost reach an equal distribution. There is a decrease of *Bifidobacterium* in subjects A and G and an increase of this genus in subjects B, D, and E in comparison to other data collection days. Subject F is missing from the figure. 

```{r plot-richness-across-day-0-7-42-with-abundance, echo=FALSE, warning=FALSE}
# Plot richness within each subject observed during each check point of the
# study. The Shannon index was used to look at species richness and evenness. 
plot_richness(phyloseq_obj,
              x = "Subject",
              color = "Day",
              measures = c("Observed", "Shannon")) +
  xlab("Subject") +
ggtitle("Observed vs Shannon Index Measure of Healthy Subjects")
```

**Figure 4:** Plot richness of each subject with comparisons of observed and Shannon index. There is an observed spread amongst each of the day samples were collected. Overlaps indicate no change or regression back to the original condition with some variation. Alpha diversity measure was used for observed and a smaller measure was seen in the Shannon index. 

```{r general-abundance-within-subjects, echo=FALSE, warning=FALSE}
# This figure allows us to draw comparisons between each subject and track 
# changes over time in regards to abundance of bacterial communities. 
melted_phyloseq %>%
  ggplot(aes(x = Subject,
             y = Abundance,
             fill = timepoints)) +
  geom_col(position = position_dodge()) +
  xlab("Subject") +
  ylab("Abundance of Bacteria") +
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1)) +
  ggtitle("Total Abundance of Bacteria Found Within Each Subject")
```

**Figure 5:** Total abundance of bacteria found in each subject, according to the date that their fecal specimen was collected and tested. Subject E showed no instance of bacteria or samples upon initial entery into the study. Subject F left before the first week due to dislike of the yogurt diet. Fluctuations of total abundance are observed and seen in various levels throughout subjects. Subject C is missing a collection sample on day 7. 

```{r boxplot-of-subject-with-treatment, echo=FALSE, warning=FALSE}
melted_phyloseq %>%
  filter(!is.na(Phylum)) %>%
  ggplot(aes(x = Treatment,
             y = Abundance,
             fill = Phylum)) +
  geom_boxplot() +
  xlab("Treatment") +
  ylab("Abundance") +
  ggtitle("Scaled Boxplot of Treatment Effects on Different Phylum Abundance") +
  theme_bw() +
  scale_y_log10() + # Use this function to observe orders of magnitude or when
  # samples are too clustured together to observe differences
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1))
```

**Figure 6:** Relative abundance of phylum presence based on the effects of both treatment types (Control and yogurt consumption) Figure shows how abundance changes over the course of the study depending on the type of treatment received. Outliers are only present within the probiotic treatment group. Greater changes of abundance averages are observed within the treatment group as well with the exception of Actinobacteria, which appeared to stay relatively neutral. 

```{r observed-plot-richness-with-alpha-diversity, echo=FALSE, warning=FALSE}
# Data frame is small enough to refer to "observed" and it demonstrates that
# averages were around the same for each time point.
plot_richness(phyloseq_obj, x = "timepoints", measures = "Observed") +
  geom_boxplot()
```

**Figure 7:** Alpha diversity measure at three distinct time points, which reveals that observed data is uniform throughout the study with noticeable outliers at each timepoint. There is a greater diversity seen at the day 42 timepoint in comparison to the other time points. The boxplot demonstrates evenness as well throughout the 3 time points. 

In my analysis I delve further into the specifics of determining changes in the GI tract of healthy individuals based on their consumption of 250 g of yogurt for 42 consecutive days. My analysis helps support the idea that regular yogurt consumption changes the structure of microbial communities, which in turn may lead to increased human fitness. Though these findings represent only a small percentage of the total relative abundance of the entire human microbiome community, it offers an explanation as to how these changes may influence greater changes. 

Fluctuation of microbial community composition is observed within each subject (Figures 1-3). *Bifidobacterium* is shown to be dominant amongst all subject microbiota, including the control (subject D), which supplies a baseline for how changes will be monitored. We observe that the presence of other bacteria are not nearly as significant, thus they are projected to play a smaller role in the GI. Each subject exhibits different levels of abundance which causes speculation over what a healthy individual is defined as. Other genuses also exhibit lower levels on day 0 of feces collection (Figure 1). On day 7 we observe that Bifidobacterium is still the dominant bacteria, yet we begin to see increasing levels of other bacteria, especially within the control group. From this time point we no longer receive samples from subject F, which drives down abundance and species diversity (Figure 2 and 3). Another missing sample that may influence abundance and diversity is a result of subject C providing an insufficient amount of fecal matter.

In contrast, species abundance of *Lactobacillus* and *Bifidobacterium* begin to reach a potential equilibrium as both groups increase in abundance (Figure 3). Out of the 8 subjects, only 5 are represented within levels of the *Genus Lactobacillus* at day 42. A striking discovery is that the control group continues to have no presence of *Lactobacillus*. The diverse nature of each subject impact is interesting as each subject was allowed to continue with their original diet and activities. In comparison to day 7, day 42 demonstrates lower abundance for the other genuses that are derived in fecal samples (Figure 3). Longitudinally, subject A and G exhibit a decrease in *Bifidobacterium* abundance, while subjects B, D, and E report an increase of this genus. However, the community composition remains diversely insignificant. 

To observe species evenness and abundance, we utilize a Shannon-Wiener index analysis versus analyzing the observed taxonomy. The Shannon plot demonstrates a decreased alpha diversity measure when scaled, which is a result of small volunteer subject pool. We find that in the beginning of the study, most of the taxa that was detected from the fecale samples were *Proteobacteria*, all of which are Gram (-) bacteria. As the timepoints continued we observe a slight decrease in Proteobacteria and begin to see an increase of gram (+) bacteria, *Firmicutes*. By the end of the analysis we see a further driving down of *Proteobacteria* and an increase in Firmicutes and Actinobacteria (Figure 6).

A summary of the changes within total abundance of bacteria supports the observation of fluctuating genus species (Figure 5). This helps to confirm that there is no congruence amongst subjects and it reveals that subject E had the greatest change in the study. Moreover, subject A displayed a steady increase in abundance and subject G demonstrated the greatest change from day 0 to day 7, then decreased drastically at day 42. Contrastly, subject B reported a sharp decrease in abundance from time point 0 to 7 then shot up again at time point 42 (Figure 5). Another unique event is seen in subject G as it had the highest abundance of bacteria noted on day 7, but then dramatically decreased on day 42. Subject E showed the most change with an almost 250 fold increase of bacteria abundance from day 7 to day 42, thus this subject faced the greatest impact out of all other subjects. 
We observe changes in phylum as well, with an average increase in communities of *Firmicutes* within all the subjects that received treatment. The only phylum that exhibited outliers was *Actinobacteria* within the treatment group. Our analysis highlights that the *phylum Proteobacteria* is very low in the control group, while *Actinobacteria* stays relatively neutral in terms of the average number between control and treatment groups (Figure 6). This plot grants a broad overview of treatment impact and shows how total treatment differs from control. 

Moreover, we shift to visualize only the findings within the observed data, which indicates that averages are the same for each time point (Figure 7). This offers a closer look into Figure 6 since we focus on just observed taxonomic values. Within this plot we detected more outliers from just the observed, thus highlighting greater diversity. Clustering together of microbial communities indicate that the communities are of similar composition. Over time we observe differentiation in the microbial community, which allows us to draw conclusions about the impact the community has on the rest of the individual. The Shannon-Wiener index provides a good estimate of ecological diversity, which allowed us to quantify the changes in gut microbial diversity. This diversity was based on terminal restriction fragments (T-RFs) and NGS universal bacteria primer data after the patients ate yogurt [@Lisko2017]. Shannon index takes into account both richness and evenness of each group. The first two weeks suggested a trend towards greater diversity, followed by a plateau and drop by the time it reached day 42. We also observe that in comparison to day 0, all test subjects had the highest Shannon Wiener diversity index except for subject A and the control subject D (Figure 4). 
Overall these findings determined that microbial communities within individual subjects diverge from the start of day 0 and become increasingly dissimilar as they continue with consecutive yogurt consumption. Variable results arise from the independent choice of primers and sequencing methods.  Thus, the three time points that the data provided act as checkpoints for each sample and their microbial communities.  The control group also reared these changes (Figure 4), which is interesting to consider since they provide insight as to how naturally the GI tract can change without probiotic consumption. 

# Discussion

The data I used for my analysis was taken from one of the first studies to report weekly monitoring of the human gut microbiota influenced by dietary yogurt rather than other concentrated applications of probiotics. Coupled with the ingestion of dietary yogurt, molecular techniques were used to better analyze the short term effects of this consumption and how it influences the GI tract. The microbiota is intimately involved in numerous aspects of normal host physiology, from nutritional status to behavior and stress response. Additionally, they can be a central or a contributing cause of many diseases, affecting both near and far organ systems. The overall balance in the composition of the gut microbial community, as well as the presence or absence of key species capable of affecting specific responses, is important in ensuring homeostasis or lack thereof at the intestinal mucosa and beyond. The mechanisms through which microbiota exerts its beneficial or detrimental influences remain largely undefined, but include elaboration of signaling molecules and recognition of bacterial epitopes by both intestinal epithelial and mucosal immune cells [@Sekirov2010].

I extensively looked at composition and diversity of the gut microbiome through the lens of healthy volunteer subjects. Although there were 8 participants to begin with, one subject was exempt due to dislike towards the intense yogurt diet and one was a control subject. The control subject D was reported to be strongly lactose intolerant and had no previous history of consuming yogurt. They were utilized in order to determine semi-quantitative patterns of microbial communities compared to other individuals that integrated yogurt consumption into their diet for a short term period [@Lisko2017]. The work done was novel as it did not involve the use of culturing techniques nor probe markers, which were traditionally used in earlier studies.

Despite the apparent benefits of yogurt consumption, studies have reported that the average yogurt consumption was only 82 g per day, which differs from the 250 g of yogurt per day that these subjects were ingesting in this study. This comes as no surprise as approximately less than 3% of the population actually consumes 250 g of yogurt or more per day [@Cormier2015]. Similar trends are found amongst North Americans as on average they consume 341 grams per week [@Wang2013]. Moreover, on a global scale, researchers have revealed that individuals from Europe and Australia tend to eat more yogurt. In these areas, low intake was less than 100 g per day, moderate was 100 to 200 g per day, and high was more than 200 g per day [@Ivey2011]. These intake trends and demographics from the literature were all considered in order to establish an ideal dose of yogurt that would increase the chances of being able to see a probiotic effect amongst the subjects. Interestingly, these subjects consumed much greater than the amount normally consumed by individuals in the US, yet is within the high end range that is found in Europe and Australia. These findings support establishing a recommended baseline of yogurt consumption and the idea that greater yogurt intake can benefit human fitness through its vast effects. 

Through molecular fingerprinting techniques and data visualization in R studio we observed that the microbial community begins to change as early as day 7 of the yogurt regimen diet (Figure 1-3). Most notably, we begin to see the abundance of *Lactobacillus species* increase as time points increase. The Shannon-Wiener index was a particularly useful measurement when determining this as it takes into account both the number of bacterial groups and abundance of each group, along with changes in rare groups that may be present within the environment [@Karlsson2010]. Fluctuations in this diversity along time points day 0, day 7, and day 42 may have resulted from the freedom that subjects were granted to continue with their regular diet and activities (Figure 4). Average values of diversity were found to be similar across the observed data (Figure 7) but fluctuated between day 0 and day 42 in the Shannon-Wiener analysis, thus demonstrating possible adaptation of the GI to the consecutive yogurt intake. 

Interestingly, among a majority of the subjects, a general trend towards greater diversity was observed, followed by a plateau and decrease in abundance and diversity on day 42 (Figure 5). This fluctuation may have been a result of rapidly changing dietary habits, thus causing a readjustment of the species found in the GI. Our findings also infer that there was no significant trend change in Bifidobacteria over the course of the study, which may have been a result of probiotics simply supporting the survival of this bacteria as opposed to being strong enough to change its composition. 
However, changes in microbial community composition are not the only things that yogurt may offer. We hypothesized that these changes are essential for increasing human fitness, specifically in reduction of stress and reduction of risk for cardiovascular diseases. The first step in determining this was to establish an actual change in composition (Figure 5), which then allowed us to determine the implications that these changes hold. By the end of the study, Lactobacillus was found to be almost equal with the abundance of *Bifidobacteria* (Figure 3), which indicates that composition was being positively affected by the yogurt consumption. Though we were unable to determine the exact species that flourished from this change of diet, the increase of Lactobacillus granted enough confidence to potential changes in the subjects fitness. For instance, we expect to see a decrease in weight of the subjects, as seen in other studies conducted. Yogurt provides substantial amounts of macronutrients and micronutrients such as calcium that is one of the main underlying assumptions to help control body weight. Recent work has been able to demonstrate that low calcium intake led to fat accumulation via an increase in blood levels of calciotropic hormones and intracellular calcium influx, thereby decreasing lipolysis and increasing lipogenesis [@Cormier2015]. Body composition measurements from a study using a similar study design showed that body fat mass was reduced in all treatments, with the greatest reduction from *Lactobacillus amylovorus* (LA) consumption. Their results support our findings that modulation of gut microbial composition from probiotic consumption may contribute to altered energy metabolism and body composition [@Omar2013]. The introduction of *Lactobacillus species* may serve as drivers in health benefits and may be potentially used as a therapeutic in future studies. 

In addition, yogurt consumption was associated with lower levels of circulating triglycerides, glucose, and lower systolic blood pressure and insulin resistance [@Wang2013]. Wang et al., observed that an average 4.7 g/d of potassium intake was related to 8.0/4.1 mm Hg lower systolic/diastolic blood pressure, 8% to 15% lower risk of cerebrovascular accident, and 6% to 11% lower risk of myocardial infarction [@Wang2013]. We expect these impacts to play an even greater role if consecutive days of consumption are increased and if other metrics are included in the analysis of subjects. These findings highlight the importance of integrating yogurt containing strains of Lactobacillus into our regular diet. Possible explanations for these health benefits may arise from the diversity that yogurt consumption creates in the GI gut, which directly have an impact on different parts of the body. 

Another health benefit that yogurt consumption catalyzes is its effect on stress-induced abdominal dysfunction, which is a popular target for probiotics. One strain of *Lactobacillus casei* that is of particular interest is Shirota. Yakult features this probiotic as its main one that helps alleviate pressure on the small intestine. Shirota has been shown to also preserve the composition of the gut microbiota and may relieve stress-associated responses of abdominal dysfunction in healthy subjects that are exposed to stressful situations [@KatoKataoka2016]. Coupled with its ability to help relieve stress indicators, Shirota also prevents infection and cancer, reduces anxiety symptoms, and suppresses the onset of physical symptoms in individuals that are exposed to academic stress. Shirota also played a role in the gut-brain axis by attenuating lipid metabolism via suppression of sympathetic nerve activity [@KatoKataoka2016]. This in turn also affects abdominal dysfunction and lowers stress levels of individuals. Another study that indicates probiotics as a method for lowering the risk of cardiovascular disease report that consumption of >100 g yogurt per day was associated with a significantly lower CCA-IMT. This relation was sufficiently robust to remain even after adjustment for their identified baseline, dietary, and lifestyle risk factors of the subjects [@Ivey2011]. They then drew comparisons with subjects with low yogurt consumption to find that those with moderate yogurt consumption had a 3.1% lower CCA-IMT. 

Future directions that this work can take is seen by the fact that values in the Shannon-Wiener index are relatively low (Figure 4) In future studies it is recommended that a larger number of volunteer subjects be used in order to establish greater distinctions of GI diversity. Making further changes in the study design, such as increasing abstention from any form of probiotics may result in a more sufficient baseline of microbial communities. By establishing this baseline future work may be able to uncover greater potential effects of yogurt on the GI microbial community. Other considerations include increasing the number of controls in the study, thus providing a more statistical robustness and comparison between treatment effects on microbial communities. The advances in modeling and analysis of gut microbiota will further our knowledge of their role in health and disease, allowing customization of existing and future therapeutic and prophylactic modalities [@Sekirov2010]. 


# Sources Cited
