---
title: "Final Report: Your Title Here"
author: "Brian Rezende"
date: "December 5, 2018"
output:
github_document: default
bibliography: references.bib
csl: bioinformatics.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Add about 2-3 pages here. Across the whole manuscript, you should cite at least 20 peer reviewed articles.

# Methods

## Study design

Add about half a page here. In this section instead of first person (I/we), use the authors of the paper you selected, since you'll just be describing what they did, based on the methods in their paper.

## Sample origin and sequencing

Add about half a page here. In this section instead of first person (I/we), use the authors of the paper you selected, since you'll just be describing what they did, based on the methods in their paper.

## Computational

These are the methods you used to do your bioinformatic analyses analyses. Should probably be between 0.5 and 1 pages. At a very minimum should include citations for DADA2 and phyloseq if you are doing an amplicon study, or other citations as appropriate.

# Results

## Subsections are ok in the results section too

```{r your-code-here}
# Add code chunks as needed for your analyses
# For most analyses, I would recommend splitting the
# intensive computational part into a seperate R script file
# and then just load your libraries and the data object here
# using the `load()` function. Ask me for clarification if this is
# unclear.
```

```{r load-packages-and-data}
# load packages as needed
library("dplyr")
library("tidyr")
library("citr")
library("knitr")
library("ggplot2")
library("seqinr")
library("mctoolsr")
library("phyloseq")
# load tree package
source("https://bioconductor.org/biocLite.R")
biocLite("ggtree")

# load data produced from analysis scripts using
# something like load("output/processed_data.Rdata")
load("output/phyloseq_obj.R")
```

```{r most-abundant-phylum-day-0}
melted_phyloseq %>%
  filter(Treatment %in%  c("Probiotic Treatment", "Probiotic Treatmen")) %>%
  

```

```{r plot-richness-across-day-0-7-42-with-abundance}

plot_richness(melted_phyloseq,
              x = Subject,
              color = Day,
              measures = c("Observed", "Shannon")) +
  xlab("Day in Study") +
ggtitle("Plot richness of subjects throughout the study")
```

```{r most-abundant-phylum-day-7}

```

```{r most-abundant-phylum-day-42}

```

```{r general-abundance-within-subjects}
melted_phyloseq %>%
ggplot(aes(x = Subject,
             y = Abundance,
             fill = Day)) +
  geom_col(position = position_dodge()) +
  xlab("Subject") +
  ylab("Abundance") +
  ggtitle("Total Abundance in Each Subject")

```
*Figure 1*: Total abundance of bacteria found in each subject, according to the date that their fecal specimen was collected and tested. Subject E showed no instance of bacteria or samples upon initially entering into the study. Subject F left before the first week due to dislike of the yogurt diet. 
```{r boxplot-of-subject-with-treatment}
metadata_in_2 %>%
ggplot(aes(x = Subject,
             y = AvgSpotLen,
             fill = Treatment)) +
   geom_boxplot() +
    facet_wrap(~Day) +
    xlab("Subject") +
    ylab("Average Spot Length") +
    ggtitle("Working title") +
    theme_bw() +
  theme(axis.text.x = element_text(angle = 90,
                         hjust = 1))

```

```{r}
metadata_in_2 %>%
  phyloseq()
```

```{r filtered-ABCF-treatment}
metadata_in_2 %>%
  filter(Subject %in% 
           c("Subject A", "Subject B", "Subject "))
```


In addition to a minimum of 5-8 figures/tables (and associated captions), you should include sufficient text in this section to describe what your findings were. Remember that in the results section you just describe what you found, but you don't interpret it - that happens in the discussion. 2-3 pages.

# Discussion

Add around 3-4 pages interpreting your results and considering future directions one might take in analyzing these data.

# Sources Cited
